Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2014-09-17 ms.tex
Some edits

\begin{abstract}
We calculate the steady-state hydrodynamic profiles of hot gas in the nuclei of a sample of early-type quiescent galaxies from the Nuker database.  We assume that mass is supplied to the circumnuclear medium (CNM) entirely by stellar winds according to the radial stellar profiles for each galaxy, while the gas is heated by a combination of shocked stellar winds and supernovae.    
Our analytic and numerical results are used to estimate the accretion rates onto the central supermassive black holes (SMBH) as a function of the SMBH mass $M_{\bullet}$ and the heating efficiency, the latter of which is related schematically to the age of the stellar population, the frequency of supernovae, and the efficacy of accretion feedback processes.  For a fixed heating efficiency, we find that the SMBH accretion Eddington factor increases with the SMBH mass, $\dot{M}/\dot{M}_{\rm edd} \propto M_{\bullet}^{\alpha}$ with $\alpha \approx 0.5, 1$ for core and cusp nuclei, respectively.  However, considering the potentially lower heating efficiency of low mass SMBHs (due to a younger stellar population and the rareity of supernovae within the stagnation radius) and their resulting higher susceptibility to cooling instability, we find a flatter relationship $\alpha \approx 0$.  Our results are used to check previous estimates of the SMBH accretion rate assuming Bondi accretion and extrapolating the X-ray measured density and temperature to the unresolved Bondi radius.  We also discuss the implications of our results for the diversity of CNM environments encountered by the jetted outflows from tidal disruption events, and for the use of an assumed $\dot{M}-M_{\bullet}$ relationship to constrain the occupation fraction of SMBHs in low mass galaxies.  
\end{abstract}

@@

The density profile of the CNM also influences the emission from stellar tidal disruption events (TDEs), such as the high energy transient {\it Swift} J1644+57 (\citealt{Levan+11}; \citealt{Bloom+11}, \citealt{Burrows+11}; \citealt{Zauderer+11}).  This event was powered by an transient relativistic jet, which produced synchrotron radio emission as the jet material was decelerated by shock interaction with the CNM of the previously quiescent SMBH (\citealt{Giannios&Metzger11}; \citealt{Zauderer+11}).  Detailed modeling of {\it Swift} J1644+57 showed that the CNM density was significantly lower than that measured surrounding Sgr A$^{\star}$ (\citealt{Metzger+12}; \citealt{Berger+12}).  However, a TDE jet injected into a denser CNM would be decelerated more rapidly, producing substantially different radio emission than in {\it Swift} J1644+57.  Variations in the properties of the CNM could thus in principle help explain why most TDEs appear to be radio quiet (e.g.~\citealt{Bower+13}; \citealt{VanVelzen+13}).

@@

Following \citet{Quataert:2004a} (see also \citealt{HolzerAxford:1970a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}), we calculate the gas density profile $\rho$ of the CNM for each galaxy within our sample by solving the equations of one-dimensional (spherically symmetric) time-dependent hydrodynamics,
\begin{align}
  &\drhodt+\frac{1}{r^2}\frac{\partial}{\partial r}\left(\rho r^2 v\right)=q \label{eq:drhodt}\\
  &\rho \left(\frac{\partial v}{\partial t} + v\frac{\partial v}{\partial r}\right) =-\dpdr-\rho \dphidr-q v \label{eq:dvdt}\\
  &\rho T\left(\frac{\partial s}{\partial t} + v\frac{\partial s}{\partial r}\right)=q\left[\ke+\kew-\gammaf \cs \right] + \nabla\cdot(\kappa \nabla T), \label{eq:dsdt}
\end{align}
where $v$, $T$, $p$ and $s$ are the radial velocity, temperature, pressure and entropy of the gas, respectively.  Here $\partial \Phi/\partial r = GM_{\rm enc}/r^{2}$ is the gravitational acceleration, where $M_{\rm enc} = M_{\star}(r) + M_{\bullet}$ is the enclosed mass.  We assume an ideal gas equation of state with $p = \rho kT/\mu m_p$ with $\mu = 0.62$ {\bf BDM: correct?} and $\gamma = 4/3$.

The source term in the continuity equation (eq.~[\ref{eq:drhodt}]),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in terms of the fraction $\eta$ of the stellar density $\rhostar$ being recycled into gas on the Hubble time $\th = 10^{10}$ years (see, e.g., \citealt{Ciotti+91}).  In Appendix \ref{app:eta} we estimate that $\eta \approx 0.11(t_{\star}/t_{\rm h})^{-1.26}$ for a Salpeter IMF, assuming that all stars are formed in a burst of star formation at a common age of $t_{\star}$ (eq.~[\ref{eq:eta}]).  We adopt $\eta=1$ as fiducial, but note that changing $\eta$ is equivalent to rescaling $\rho$ by the same factor.\footnote{This is {\it not} true once cooling is included, since the cooling rate is proportional to density {\it squared}.}

Source terms $\propto q$ also appear in the momentum and entropy equations (eq.~[\ref{eq:dvdt}], [\ref{eq:dsdt}]) due to the fact that the isotropic mass injection in the rest frame of the SMBH represents a source of momentum and energy upon being incorporated into the mean flow (see also below).  The second source term in equation (\ref{eq:dsdt}) accounts for heating transportation by electron conductivity, where the conductivity $\kappa = \kappa_0 T^{5/2}$ is assumed to equal the classical ``Spitzer" value (\citealt{Spitzer62}); although even a weak magnetic field could in principle strongly suppress the conductivity perpendicular to the field, the radially-decreasing temperature profiles of interest are susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field structure (\citealt{Parrish&Stone07}).

@@

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved for each galaxy using a standard Runge-Kutta integration for different assumed values of $v_{\rm w,0}$ spanning the physically expected range.  Although we are most interested in the steady-state CNM structure, the time-dependent equations are solved to avoid numerical issues that arise near the critical sonic points, of which there are up to two because of the supersonic inflow on small radial scales and a possible supersonic outflow on large scales.

We do not include radiative cooling in the entropy equation because across much of the parameter space radiative cooling is negligible compared to other sources of cooling (e.g., advection).  Radiative cooling cannot be neglected in some cases (especially if $\eta$ is high), but the onset of a thermal instability when radiative cooling dominates precludes obtaining a steady-state solution anyways.  Our steady state solutions neglecting radiation can nevertheless be used to ascertain over what parameter space cooling instabilities are likely to develop ($\S\ref{sec:cooling}$).

The accuracy of our numerical solutions is assessed by checking the level of conservation of mass across the grid, as well as the integral constraint on the Bernoulli parameter.  Mass and Bernoulli are conserved to $\lesssim 10$ per cent for all solutions presented in this paper.  {\bf BDM: more details...}

@@

First, by demanding continuity of the entropy derivative at the stagnation radius, the temperature at the stagnation radius is given by (eq.~[\ref{eq:Tanalytic}])
\begin{align}
T \simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p v_{w}^{2}}{2} \approx 4 \times 10^6\,v_{500}^2 \, \chi   \,\,{\rm K} 
\end{align}
{\bf BDM: i picked $\mu = 0.62$ and stick to it.}

Next, by demanding hydrostatic equilibrium at the stagnation point, an expression for the stagnation radius can be derived (eq.~[\ref{eq:rstag}]),
\begin{align}
  \x=\frac{1}{\zeta^2 n}\left[ (1+\Mstar(\rs)/\Mbh)\left(A-\frac{n}{2}\right)-B\right],
  \label{eq:stag_analytic}
\end{align}
where $\zeta \equiv v_{w,0}/\sigma$ {\bf BDM: note new definition}, $r_{\rm soi} = GM_{\bullet}/\sigma^{2}$ is the radius of the SMBH sphere of influence (soi), and A/B are constants that depend on the Nuker parameter $\Gamma$ and the adiabatic index of the gas.

If the SMBH dominates the potential at the stagnation radius ($M_{\bullet} \gg M_{\star}$; $v_{w,0} \gg 200 M_{\bullet,8}^{0.2}$ km s$^{-1}$) then the relationship between $\rs$ and $v_{w}$ is simplified greatly:
\begin{align}
  &\rs=\frac{7}{2}\frac{G \Mbh}{n \vwO^2} \simeq 6 \, \pc \,\, \Mbheight v_{500}^{-2},
  \label{eq:stag_simple}
\end{align}
where we have assumed $n = 1$ {\bf BDM: true?  should there be a $\chi$ in this?}.  Note that this expression is similar to that obtained by \citet{Volonteri+11} on heuristic grounds (their eq.~[6]).

The SMBH accretion rate is simply given by the integrated mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc(\rs)}{\th} \nonumber \\
&\approx&
  \begin{cases}
    1.6 \times 10^{-3}  M_{\bullet,8}^{1.9}
    v_{500}^{-4} \chi^{-2} \eta \Msun \, \pyear& \text{core} \\
    3.4 \times 10^{-3} \Mbheight^{1.4}
    v_{500}^{-2} \chi^{-1} \eta \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where the numerical estimates again assume that $M_{\bullet} \gg M_{\star}$ at the stagnation radius.  Note that equation (\ref{eq:mdot_analytic}) predicts that $\dot{M}$ should increase with BH mass at fixed $v_{500}$ but that $\dot{M}$ is extremely sensitive to the assumed heating rate $\propto v_{500}$.

@@

Rewriting equation (\ref{eq:mdot_gas}) as
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{t_{\ff}} =\frac{4 \pi}{3} \rs^2 \rho(\rs) v_{\ff}(\rs)
\end{align}
we note the qualitative similarity between this expression and the Bondi formula (eq.~[\ref{eq:bondi}]) with the exception that $\rb$ is replaced by $\rs$.  Using equation (\ref{eq:stag_simple})
\begin{align}
  \rs=\frac{7}{2}\frac{G \Mbh}{\vwO^2} \approx \frac{14}{3}\frac{GM_{\bullet}}{c_{s}^{2}}\frac{v_{w}^{2}}{v_{w,0}^{2}},
  \label{eq:rs_simple}
\end{align}
where in the second line we have used the fact that $v_{w}^{2}/2 = c_s^2/(\gamma-1)$ at $\rs$ ({\bf BDM: Check - as this depends on whether eq 10 is correct}).  Thus, provided that a stagnation radius exists which is interior to the sphere of influence radius, the accretion rate predicted by our model is quite close to $\dot{M}_{\rm B}$.

@@

Figure~\ref{fig:profiles} shows three representative profiles of the density $\rho(r)$, temperature $T(r)$, and radial velocity $v(r)$, as well as the cumulative X-ray luminosity.  As expected, the gas density increases towards the SMBH, with a characteristic radial slope $\rho \propto r^{n}$ with $n \approx 1$ near the stagnation radius (marked as a square in Fig.~\ref{fig:profiles}; {\bf BDM:check!?}).  On very small scales the gas density profile appropaches the behavior $\rho \propto r^{-3/2}$ predicted for Bondi accretion.  The temperature profile is relatively flat at large radii, but increases as $\propto 1/r$ interior to the sphere of influence, both behaviors expected for virialized gas in near hydrostatic equilibrium.  The inwardly directed velocity increases towards the hole approaching the expected free-fall rate $v \propto v_{\rm ff}\propto r^{-1/2}$ interior to the sonic point, and then increases outwards towards large radii. {\bf BDM: I made this up - in general we need more explanation of what the plots are actually showing, with some physical explanation.}

Figure~\ref{fig:stag} (top panel) shows numerical results for the stagnation radius $r_{\rm s}$ where the radial velocity goes to zero (expressed in ratio to the radius of the SMBH sphere of influence $r_{\rm soi}$), with different colors representing different values of $v_{w,0}$.  Cusp and core galaxies are marked with square and triangles, respectively.  The solid line is a power law fit through the cusp galaxies, while the dotted line is a power law fit through the core galaxies.  {\bf BDM: I see no lines!}

@@

\subsubsection{Millisecond Pulsars} The spin down power of millisecond pulsars (MSPs) could in principle be an important heating source.  If the number of MSPs per unit stellar mass is $n_{\rm msp}$ and each contributes an average spin-down luminosity $L_{\rm sd}$, then the resulting heating per unit volume $e \approx L_{\rm sd}n_{\rm msp}\epsilon$ corresponds to an effective heating rate (eq.~ [\ref{eq:vw_eff}])
  \begin{align}
    \vwO=\sqrt{\frac{2 t_h n_{\rm msp} L_{\rm sd}}{\eta}},
  \end{align}
where $\eta < 1$ is the thermalization efficiency of the wind.  The $\sim 30,000$ estimated MSPs in the Milky Way of stellar mass $6\times 10^{10}M_{\odot}$ (\citealt{Lorimer13}) corresponds to $n_{\rm msp}\simeq3 \times 10^{-40} $ MSPs g$^{-1}$.  If we assume an average spin-down luminosity $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic of Fermi pulsars, then we find that $\vwO\simeq 120\epsilon (\eta/0.1)^{-1/2}$ km s$^{-1}$.  Thus, assuming a high thermalization efficiency $\epsilon \sim 1$, MSP heating can be comparable to main sequence heating.  However, much lower thermalization efficiencies of $\epsilon \lesssim 0.1$ are inferred based on modeling the insterstellar mediums of globular clusters \citep{NaimanSoares-Furtado+:2013a}, while the binaries giving rise to through accretion MSPs may be disocciated by stellar interactions in the dense nuclear cluster.  
\subsubsection{Type Ia Supernovae}

Energy injected by Type Ia supernovae (SNe Ia) represents a potentially important source of heating, which unlike core collapse SNe is present even in an evolved stellar population.  Assuming that each Ia injects a thermal energy $E_{\rm Ia}$ into the instellar medium, and that the SNe rate (per stellar mass) is given by $R_{\rm Ia}$, then the resulting volumetric heating rate $E_{\rm Ia}R_{\rm Ia}$ produces an effective wind heating parameter (eq.~[\ref{eq:vw_eff}])
\begin{align}
    \vwO=\sqrt{\frac{2 t_h R_{\rm Ia} E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align}
The thermal energy injected by a Ia SN $E_{\rm Ia} \simeq \epsilon_{\rm Ia} 10^{51}$ ergs depends on the efficiency $\epsilon$ with which the blast wave energy is converted into bulk or turbulent motion instead of being lost to radiation.  \cite{Thornton+98} estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$, depending weakly on surrounding density, but \citet{Sharma+14} argues that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if the SNe occur in a hot dilute medium, as may characterize the CNM.

@@

In the limit of an old stellar population of age $t_{\star} \sim t_{\rm h}$ the Ia rate just becomes $R_{\rm Ia} \sim 2\times 10^{-55} $ g$^{-1}$ s$^{-1}$ (e.g.~\citealt{Scannapieco&Bildsten05}).  Likewise, as already discussed, the mass loss rate $\eta \sim 0.1-1$ also depends on the stellar age (eq.~[\ref{eq:eta}]).  Thus, from equation (\ref{eq:vw_eff}) we estimate $v_{w,0} \approx 350\epsilon_{\rm Ia}^{0.5}(\eta/0.1)^{-0.5}$ for an old stellar population.  For a younger stellar population the increasing rate of Ia SN is compensated by the larger stellar mass loss rate, resulting in a value of $v_{w,0}$ that is roughly independent of stellar age.  The high value of $v_{w,0}$ implies that Ia SNe represent a potentially dominant source of heating as compared to ms pulsars or main sequence winds.

However, SNe can only be approximated as constant source of heating throughout space and time if the rate of SNe is rapid compared to the inflow time at the radius of interest (\citealt{ShcherbakovWong+:2014a}).  To estimate when this approximation is valid, we estimate the ``Ia radius" 
\begin{eqnarray}
r_{\rm Ia} \sim \left(\frac{G}{R_{\rm Ia}\sigma}\right)^{1/2} \approx 40\left(\frac{\sigma}{200\,{\rm km\,s^{-1}}}\right)^{-1/2} \; {\rm pc} \sim 43 M_{\bullet,8}^{0.1}\,{\rm pc}
\label{eq:rIa}
\end{eqnarray}
as the location exterior of which the time interval between subsequent supernovae $\tau_{\rm Ia} \sim 1/M_{\rm enc}R_{\rm Ia} \sim G/r\sigma^{2}R_{\rm Ia}$ exceeds the local dynamical timescale $t_{\rm dyn} \sim r/\sigma$, where we again adopt the Ia rate for an old stellar population and in the final equality we estimate the velocity dispersion using the $M_{\bullet}-\sigma$ relationship (eq.~[\ref{eq:Msigma}]).

>>>>

\begin{abstract}
  We calculate the steady-state hydrodynamic profiles of hot gas in
  the nuclei of a sample of early-type quiescent galaxies from the
  Nuker database.  We assume that mass is supplied to the
  circumnuclear medium (CNM) entirely by stellar winds according to
  the radial stellar profiles for each galaxy, while the gas is heated
  by a combination of shocked stellar winds and supernovae.  Our
  analytic and numerical results are used to estimate the accretion
  rates onto the central supermassive black holes (SMBH) as a function
  of the SMBH mass $M_{\bullet}$ and the heating efficiency, the
  latter of which is related schematically to the age of the stellar
  population, the frequency of supernovae, and the efficacy of
  accretion feedback processes.  For a fixed heating efficiency, we
  find that the SMBH accretion Eddington factor increases with the
  SMBH mass, $\dot{M}/\dot{M}_{\rm edd} \propto M_{\bullet}^{\alpha}$
  with $\alpha \approx 0.5, 1$ for core and cusp nuclei, respectively.
  % However, considering the potentially lower heating efficiency of low
  % mass SMBHs (due to a younger stellar population and the rareity of
  % supernovae within the stagnation radius) and their resulting higher
  % susceptibility to cooling instability, we find a flatter
  % relationship $\alpha \approx 0$. 
  Our results are used to check previous estimates of the SMBH
  accretion rate assuming Bondi accretion and extrapolating the X-ray
  measured density and temperature to the unresolved Bondi radius.  We
  also discuss the implications of our results for the diversity of
  CNM environments encountered by the jetted outflows from tidal
  disruption events, and for the use of an assumed
  $\dot{M}-M_{\bullet}$ relationship to constrain the occupation
  fraction of SMBHs in low mass galaxies.
\end{abstract}

@@

The density profile of the CNM also influences the emission from
stellar tidal disruption events (TDEs), such as the high energy
transient {\it Swift} J1644+57 (\citealt{Levan+11};
\citealt{Bloom+11}, \citealt{Burrows+11}; \citealt{Zauderer+11}).
This event was powered by a transient relativistic jet, which produced
synchrotron radio emission as the jet material was decelerated by
shock interaction with the CNM of the previously quiescent SMBH
(\citealt{Giannios&Metzger11}; \citealt{Zauderer+11}).  Detailed
modeling of {\it Swift} J1644+57 showed that the CNM density was
significantly lower than that measured surrounding Sgr A$^{\star}$
(\citealt{Metzger+12}; \citealt{Berger+12}).  However, a TDE jet
injected into a denser CNM would be decelerated more rapidly,
producing substantially different radio emission than in {\it Swift}
J1644+57.  Variations in the properties of the CNM could thus in
principle help explain why most TDEs appear to be radio quiet
(e.g.~\citealt{Bower+13}; \citealt{VanVelzen+13}).

@@

Following \citet{Quataert:2004a} (see also \citealt{HolzerAxford:1970a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}), we calculate the gas density profile $\rho$ of the CNM for each galaxy within our sample by solving the equations of one-dimensional (spherically symmetric) time-dependent hydrodynamics,
\begin{align}
  &\drhodt+\frac{1}{r^2}\frac{\partial}{\partial r}\left(\rho r^2 v\right)=q \label{eq:drhodt}\\
  &\rho \left(\frac{\partial v}{\partial t} + v\frac{\partial v}{\partial r}\right) =-\dpdr-\rho \dphidr-q v \label{eq:dvdt}\\
  &\rho T\left(\frac{\partial s}{\partial t} + v\frac{\partial
      s}{\partial r}\right)=q\left[\ke+\kew-\gammaf \cs \right] 
  %+\nabla\cdot(\kappa \nabla T)
, 
\label{eq:dsdt}
\end{align}
where $v$, $T$, $p$ and $s$ are the radial velocity, temperature,
pressure and specific entropy of the gas, respectively.  Here $\partial
\Phi/\partial r = GM_{\rm enc}/r^{2}$ is the gravitational
acceleration, where $M_{\rm enc} = M_{\star}(r) + \Mbh$ is the
enclosed mass.  We assume an ideal gas equation of state with $p =
\rho kT/\mu m_p$ with $\mu = 1$  and $\gamma = 5/3$.

The source term in the continuity equation (equation~\ref{eq:drhodt}),
%Should equations be referenced as equation of equation
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 10^{10}$ years (see, e.g.,
\citealt{Ciotti+91}).  In Appendix \ref{app:eta} we estimate that
$\eta \approx 0.11(t_{\star}/t_{\rm h})^{-1.26}$ for a Salpeter IMF,
assuming that all stars are formed in a burst of star formation at a
common age of $t_{\star}$ (equation~\ref{eq:eta}).  We adopt
$\eta=1$ as fiducial, but note that changing $\eta$ is equivalent to
rescaling $\rho$ by the same factor.\footnote{This is {\it not} true
  if cooling is included, since the cooling rate is proportional to
  density {\it squared}.}

Source terms $\propto q$ also appear in the momentum and entropy
equations (equation~\ref{eq:dvdt} and \ref{eq:dsdt}) due to the fact
that the isotropic mass injection in the rest frame of the SMBH
represents a source of momentum and energy upon being incorporated
into the mean flow (see also below).  
%No net momentum is in the winds, but mass is being added which adds a
%term in the velocity equation...

% The second source term in
% equation (\ref{eq:dsdt}) accounts for heating transportation by
% electron conductivity, where the conductivity $\kappa = \kappa_0
% T^{5/2}$ is assumed to equal the classical ``Spitzer" value
% (\citealt{Spitzer62}); although even a weak magnetic field could in
% principle strongly suppress the conductivity perpendicular to the
% field, the radially-decreasing temperature profiles of interest are
% susceptible to the magneto-thermal instability (\citealt{Balbus01}),
% the non-linear evolution of which is believed to result in a
% radially-directed field structure (\citealt{Parrish&Stone07}).

@@

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability. This scheme is
described in \citealt{Brandenburg:2003a}.  We solve the equations for
different choices of $v_{\rm w,0}$ spanning the physically expected
range.  Although we are most interested in the steady-state CNM
structure, the time-dependent equations are solved to avoid numerical
issues that arise near the critical sonic points, of which there are
up to two because of the supersonic inflow on small radial scales and
a possible supersonic outflow on large scales.

We do not include radiative cooling in the entropy equation because
across much of the parameter space radiative cooling is negligible
compared to other sources of cooling (e.g., advection).  Radiative
cooling cannot be neglected in some cases (especially if $\vwO$ is low
or $\eta$ is high), but the onset of a thermal instability when
radiative cooling dominates precludes obtaining a steady-state
solution anyways.  Our steady state solutions neglecting radiation can
nevertheless be used to ascertain over what parameter space cooling
instabilities are likely to develop ($\S\ref{sec:cooling}$).
%Dependence of the heating to cooling ratio has a far steeper
%dependence on vwO than on eta...

The accuracy of our numerical solutions is assessed by checking the
level of conservation of mass across the grid, as well as the integral
constraint on the energy.  Mass and energy are conserved to $\lesssim
40$ per cent for all solutions presented in this
paper. % {\bf BDM: more details...}

@@

First, by demanding continuity of the entropy derivative at the stagnation radius, the temperature at the stagnation radius is given by (equation~[\ref{eq:Tanalytic}])
\begin{align}
T \simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p v_{w}^{2}}{2} \approx 4 \times 10^6\,v_{500}^2 \, \chi   \,\,{\rm K} 
\end{align}
{\bf BDM: i picked $\mu = 0.62$ and stick to it.}

Next, by demanding hydrostatic equilibrium at the stagnation point, an expression for the stagnation radius can be derived (equation~[\ref{eq:rstag}]),
\begin{align}
  \x=\frac{1}{\zeta^2 n}\left[ (1+\Mstar(\rs)/\Mbh)\left(A-\frac{n}{2}\right)-B\right],
  \label{eq:stag_analytic}
\end{align}
where $\zeta \equiv v_{w,0}/\sigma$ {\bf BDM: note new definition}, $r_{\rm soi} = GM_{\bullet}/\sigma^{2}$ is the radius of the SMBH sphere of influence (soi), and A/B are constants that depend on the Nuker parameter $\Gamma$ and the adiabatic index of the gas.

If the SMBH dominates the potential at the stagnation radius
($M_{\bullet} \gg M_{\star}$; $v_{w,0} \gg 200 M_{\bullet,8}^{0.2}$ km
s$^{-1}$) then the relationship between $\rs$ and $v_{w}$ is
simplified greatly:
\begin{align}
  &\rs=\frac{7}{2}\frac{G \Mbh}{n \vwO^2} \simeq 6 \, \pc \,\, \Mbheight v_{500}^{-2},
  \label{eq:stag_simple}
\end{align}
where we have assumed $n = 1$ {\bf BDM: true?  should there be a
  $\chi$ in this?}.  Note that this expression is similar to that
obtained by \citet{Volonteri+11} on heuristic grounds (their equation~[6]).

The SMBH accretion rate is simply given by the integrated mass loss rate interior to the stagnation radius (equation~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc(\rs)}{\th} \nonumber \\
&\approx&
  \begin{cases}
    1.6 \times 10^{-3}  M_{\bullet,8}^{1.9}
    v_{500}^{-4} \chi^{-2} \eta \Msun \, \pyear& \text{core} \\
    3.4 \times 10^{-3} \Mbheight^{1.4}
    v_{500}^{-2} \chi^{-1} \eta \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where the numerical estimates again assume that $M_{\bullet} \gg M_{\star}$ at the stagnation radius.  Note that equation (\ref{eq:mdot_analytic}) predicts that $\dot{M}$ should increase with BH mass at fixed $v_{500}$ but that $\dot{M}$ is extremely sensitive to the assumed heating rate $\propto v_{500}$.

@@

Rewriting equation (\ref{eq:mdot_gas}) as
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{t_{\ff}} =\frac{4 \pi}{3} \rs^2 \rho(\rs) v_{\ff}(\rs)
\end{align}
we note the qualitative similarity between this expression and the Bondi formula (equation~[\ref{eq:bondi}]) with the exception that $\rb$ is replaced by $\rs$.  Using equation (\ref{eq:stag_simple})
\begin{align}
  \rs=\frac{7}{2}\frac{G \Mbh}{\vwO^2} \approx \frac{14}{3}\frac{GM_{\bullet}}{c_{s}^{2}}\frac{v_{w}^{2}}{v_{w,0}^{2}},
  \label{eq:rs_simple}
\end{align}
where in the second line we have used the fact that $v_{w}^{2}/2 =
c_s^2/(\gamma-1)$ at $\rs$ ({\bf BDM: Check - as this depends on
  whether eq 10 is correct}).  Thus, provided that a stagnation radius
exists which is interior to the sphere of influence radius, the
accretion rate predicted by our model is quite close to $\dot{M}_{\rm
  B}$.

@@

Figure~\ref{fig:profiles} shows three representative profiles of the
density $\rho(r)$, temperature $T(r)$, and radial velocity $v(r)$, as
well as the cumulative X-ray luminosity.  As expected, the gas density
increases towards the SMBH, with a characteristic radial slope $\rho
\propto r^{n}$ with $n \approx 1$ near the stagnation radius (marked
as a square in Fig.~\ref{fig:profiles}. The gas density The
temperature profile is relatively flat at large radii, but increases
as $\propto 1/r$ interior to the sphere of influence, both behaviors
expected for virialized gas in near hydrostatic equilibrium.  The
inwardly directed velocity increases towards the hole approaching the
expected free-fall rate $v \propto v_{\rm ff}\propto r^{-1/2}$
interior to the sonic point, and then increases outwards towards large
radii. {\bf BDM: I made this up - in general we need more explanation
  of what the plots are actually showing, with some physical
  explanation.}

Figure~\ref{fig:stag} (top panel) shows numerical results for the
stagnation radius $r_{\rm s}$ where the radial velocity goes to zero
(expressed in ratio to the radius of the SMBH sphere of influence
$r_{\rm soi}$), with different colors representing different values of
$v_{w,0}$.  Cusp and core galaxies are marked with square and
triangles, respectively.  
%The solid line is a power law fit through the cusp galaxies, while the dotted line is a power law fit through the core galaxies.  {\bf BDM: I see no lines!}

@@

\subsubsection{Millisecond Pulsars} The spin down power of millisecond pulsars (MSPs) could in principle be an important heating source.  If the number of MSPs per unit stellar mass is $n_{\rm msp}$ and each contributes an average spin-down luminosity $L_{\rm sd}$, then the resulting heating per unit volume $e \approx L_{\rm sd}n_{\rm msp}\epsilon$ corresponds to an effective heating rate (equation~ [\ref{eq:vw_eff}])
  \begin{align}
    \vwO=\sqrt{\frac{2 t_h n_{\rm msp} L_{\rm sd}}{\eta}},
  \end{align}
where $\eta < 1$ is the thermalization efficiency of the wind.  The $\sim 30,000$ estimated MSPs in the Milky Way of stellar mass $6\times 10^{10}M_{\odot}$ (\citealt{Lorimer13}) corresponds to $n_{\rm msp}\simeq3 \times 10^{-40} $ MSPs g$^{-1}$.  If we assume an average spin-down luminosity $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic of Fermi pulsars, then we find that $\vwO\simeq 120\epsilon (\eta/0.1)^{-1/2}$ km s$^{-1}$.  Thus, assuming a high thermalization efficiency $\epsilon \sim 1$, MSP heating can be comparable to main sequence heating.  However, much lower thermalization efficiencies of $\epsilon \lesssim 0.1$ are inferred based on modeling the insterstellar mediums of globular clusters \citep{NaimanSoares-Furtado+:2013a}, while the binaries giving rise to through accretion MSPs may be disocciated by stellar interactions in the dense nuclear cluster.  
\subsubsection{Type Ia Supernovae}

Energy injected by Type Ia supernovae (SNe Ia) represents a potentially important source of heating, which unlike core collapse SNe is present even in an evolved stellar population.  Assuming that each Ia injects a thermal energy $E_{\rm Ia}$ into the instellar medium, and that the SNe rate (per stellar mass) is given by $R_{\rm Ia}$, then the resulting volumetric heating rate $E_{\rm Ia}R_{\rm Ia}$ produces an effective wind heating parameter (equation~[\ref{eq:vw_eff}])
\begin{align}
    \vwO=\sqrt{\frac{2 t_h R_{\rm Ia} E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align}
The thermal energy injected by a Ia SN $E_{\rm Ia} \simeq \epsilon_{\rm Ia} 10^{51}$ ergs depends on the efficiency $\epsilon$ with which the blast wave energy is converted into bulk or turbulent motion instead of being lost to radiation.  \cite{Thornton+98} estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$, depending weakly on surrounding density, but \citet{Sharma+14} argues that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if the SNe occur in a hot dilute medium, as may characterize the CNM.

@@

In the limit of an old stellar population of age $t_{\star} \sim t_{\rm h}$ the Ia rate just becomes $R_{\rm Ia} \sim 2\times 10^{-55} $ g$^{-1}$ s$^{-1}$ (e.g.~\citealt{Scannapieco&Bildsten05}).  Likewise, as already discussed, the mass loss rate $\eta \sim 0.1-1$ also depends on the stellar age (equation~[\ref{eq:eta}]).  Thus, from equation (\ref{eq:vw_eff}) we estimate $v_{w,0} \approx 350\epsilon_{\rm Ia}^{0.5}(\eta/0.1)^{-0.5}$ for an old stellar population.  For a younger stellar population the increasing rate of Ia SN is compensated by the larger stellar mass loss rate, resulting in a value of $v_{w,0}$ that is roughly independent of stellar age.  The high value of $v_{w,0}$ implies that Ia SNe represent a potentially dominant source of heating as compared to ms pulsars or main sequence winds.

However, SNe can only be approximated as constant source of heating throughout space and time if the rate of SNe is rapid compared to the inflow time at the radius of interest (\citealt{ShcherbakovWong+:2014a}).  To estimate when this approximation is valid, we estimate the ``Ia radius" 
\begin{eqnarray}
r_{\rm Ia} \sim \left(\frac{G}{R_{\rm Ia}\sigma}\right)^{1/2} \approx 40\left(\frac{\sigma}{200\,{\rm km\,s^{-1}}}\right)^{-1/2} \; {\rm pc} \sim 43 M_{\bullet,8}^{0.1}\,{\rm pc}
\label{eq:rIa}
\end{eqnarray}
as the location exterior of which the time interval between subsequent supernovae $\tau_{\rm Ia} \sim 1/M_{\rm enc}R_{\rm Ia} \sim G/r\sigma^{2}R_{\rm Ia}$ exceeds the local dynamical timescale $t_{\rm dyn} \sim r/\sigma$, where we again adopt the Ia rate for an old stellar population and in the final equality we estimate the velocity dispersion using the $M_{\bullet}-\sigma$ relationship (equation~[\ref{eq:Msigma}]).

Note that the stagnation radius scales approximately linearly with $M_{\bullet}$ (equation~[\ref{eq:stag_simple}]), while $\rIa$ is roughly indpendent of SMBH mass.  For small $\Mbh$, the stagnation radius thus lies well inside $\rIa$ and hence Ia SNe will not contribute appreciably to the heating rate.  On the other hand, for massive SMBHs $M_{\bullet} \gtrsim 10^8 \Msun$, SNe Ia  represent an important, and possibly dominant, contribution to the overall heating.
  % Thus, for $\Mbh=10^6 \Msun$, we may have $\vwO\simeq 100$ km/s with
  % heating mostly coming from main sequence winds. On the other hand,
  % for $\Mbh \gsim 10^8 \Msun$, we may have $\vwO\simeq 500$ km/s, with
  % the main contribution coming from type Ia SNe. These two regimes are
  % illustrated in Figure~\ref{fig:vw_eff}.
  % \end{enumerate}
